Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood

### 2025-09-19 paper/paper.Rmd
attempts to knit paper and supp

# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
author:
  - name: Michael C. Frank
    affiliation: a, 1
  - name: Virginia A. Marchman
    affiliation: a
  - name: Claire Augusta Bergey
    affiliation: b
  - name: Veronica Boyce
    affiliation: a
  - name: Mika Braginsky
    affiliation: a
  - name: George Kachergis
    affiliation: a
  - name: Jess Mankewitz
    affiliation: c
  - name: Stephan Meylan
    affiliation: d
  - name: Ben Prystawski
    affiliation: a
  - name: Nilam Ram
    affiliation: a, e
  - name: Robert Z. Sparks
    affiliation: a
  - name: Adrian Steffan
    affiliation: f
  - name: Alvin Wei Ming Tan
    affiliation: a
  - name: Martin Zettersten
    affiliation: g
address:
  - code: a
    address: Department of Psychology, Stanford University, Stanford, CA, USA
  - code: b
    address: Department of Linguistics, Stanford University, Stanford, CA, USA
  - code: c
    address: Department of Psychology, University of Wisconsin, Madison, WI, USA
  - code: d
    address: Department of Linguistics, University of California, Berkeley, CA, USA
  - code: e
    address: Department of Communication, Stanford University, Stanford, CA, USA
  - code: f
    address: Department of Psychology, Ludwig Maximilian University, Munich, Germany
  - code: g
    address: Department of Cognitive Science, University of California, San Diego, CA, USA

@@

## change to true to add optional line numbering
lineno: false

output: papaja::apa6_pdf

@@

\dropcap{C}hildren acquiring a language are learning a body of knowledge -- a set of words and the ways they are combined -- but they are also learning to deploy this knowledge in the myriad complex, noisy, and fast-moving environments in which language is used. As children enter their second year, language explodes onto the scene; both vocabulary and grammatical abilities grow rapidly and in tandem [@bates1994developmental;@frank2021]. This growth in knowledge is also accompanied by changes in language processing efficiency: children become quicker and more accurate in recognizing words and matching them with their referents [@fernald1998;@peter2019;@bergelson2020comprehension].

@@

```{r age_regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age. Observed variables are notated as squares and latent variables are notated as circles Factor loadings and regression coefficients are shown with straight, solid lines; covariances are shown with dashed lines; residual variances are shown as solid circular connections. Stars show conventional levels of statistical significance, e.g. * indicates p < .05, ** indicates p < .01, and *** indicates p < .001. Covariances reflect age-residualized correlations between variables.", fig.height = 3}

@@

Figure \ref{fig:age_regression} shows a regression model fit to this confirmatory factor analysis, with log age predicting each latent variable. This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age). The non-age related variance of all three latent factors was significantly related to that of the other factors, with speed and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$). This model supports the idea that variation in speed and accuracy of word recognition is related to individual differences in parent-reported vocabulary beyond the effects of age. Further, the broader set of analyses support a factor structure in which speed and accuracy (and their associated variabilities) are related but distinct aspects of word recognition, rather than being measures of one single construct. These analyses treat all data as between person, however, rather than modeling change in these factors within individuals.

@@

<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
\showmatmethods
\showacknow
\pnasbreak

\newpage

>>>>

shorttitle: Continuous developmental changes
# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
author:
  - name: Michael C. Frank
    affiliation: a, 1
  - name: Virginia A. Marchman
    affiliation: a
  - name: Claire Augusta Bergey
    affiliation: b
  - name: Veronica Boyce
    affiliation: a
  - name: Mika Braginsky
    affiliation: a
  - name: George Kachergis
    affiliation: a
  - name: Jess Mankewitz
    affiliation: c
  - name: Stephan Meylan
    affiliation: d
  - name: Ben Prystawski
    affiliation: a
  - name: Nilam Ram
    affiliation: a, e
  - name: Robert Z. Sparks
    affiliation: a
  - name: Adrian Steffan
    affiliation: f
  - name: Alvin Wei Ming Tan
    affiliation: a
  - name: Martin Zettersten
    affiliation: g
address:
  - code: a
    address: Department of Psychology, Stanford University, Stanford, CA, USA
  - code: b
    address: Department of Linguistics, Stanford University, Stanford, CA, USA
  - code: c
    address: Department of Psychology, University of Wisconsin, Madison, WI, USA
  - code: d
    address: Department of Linguistics, University of California, Berkeley, CA, USA
  - code: e
    address: Department of Communication, Stanford University, Stanford, CA, USA
  - code: f
    address: Department of Psychology, Ludwig Maximilian University, Munich, Germany
  - code: g
    address: Department of Cognitive Science, University of California, San Diego, CA, USA

@@

## change to true to add optional line numbering
lineno: false
floatsintext: yes
output: papaja::apa6_pdf

@@

Children acquiring a language are learning a body of knowledge -- a set of words and the ways they are combined -- but they are also learning to deploy this knowledge in the myriad complex, noisy, and fast-moving environments in which language is used. As children enter their second year, language explodes onto the scene; both vocabulary and grammatical abilities grow rapidly and in tandem [@bates1994developmental;@frank2021]. This growth in knowledge is also accompanied by changes in language processing efficiency: children become quicker and more accurate in recognizing words and matching them with their referents [@fernald1998;@peter2019;@bergelson2020comprehension].

@@

```{r age-regression, fig.env="figure*", fig.cap="Structural equation model showing the three-factor factor analysis with a regression of each latent variable on the logarithm of age. Observed variables are notated as squares and latent variables are notated as circles Factor loadings and regression coefficients are shown with straight, solid lines; covariances are shown with dashed lines; residual variances are shown as solid circular connections. Stars show conventional levels of statistical significance, e.g. * indicates p < .05, ** indicates p < .01, and *** indicates p < .001. Covariances reflect age-residualized correlations between variables.", fig.height = 3}

@@

Figure \ref{fig:age-regression} shows a regression model fit to this confirmatory factor analysis, with log age predicting each latent variable. This regression model allows interpretation of the covariances between latent factors as partial correlations (controlling for age). The non-age related variance of all three latent factors was significantly related to that of the other factors, with speed and accuracy showing strong negative covariance ($\beta=$ `r round(fit3_age_summary$pe[21,"est"],3)`, SE = `r round(fit3_age_summary$pe[21,"se"],3)`, $p < .0001$) and weaker but significant covariation between RT and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[20,"est"],3)`, SE = `r round(fit3_age_summary$pe[20,"se"],3)`, $p < .0001$) and accuracy and vocabulary ($\beta=$ `r round(fit3_age_summary$pe[19,"est"],3)`, SE = `r round(fit3_age_summary$pe[19,"se"],3)`, $p < .0001$). This model supports the idea that variation in speed and accuracy of word recognition is related to individual differences in parent-reported vocabulary beyond the effects of age. Further, the broader set of analyses support a factor structure in which speed and accuracy (and their associated variabilities) are related but distinct aspects of word recognition, rather than being measures of one single construct. These analyses treat all data as between person, however, rather than modeling change in these factors within individuals.

@@

<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
<!--\showmatmethods
\showacknow
\pnasbreak

\newpage -->
\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

### 2025-09-19 paper/supplemental_information.Rmd
attempts to knit paper and supp

```{r}
#load data using Creel (2024) analysis scripts
# To run this script, you need to first pull in two datasets from the Creel (2024) OSF repository (https://osf.io/wszd5/):
#allfamiliars.csv and allnovels.csv
library(osfr)
creel_proj <- osf_retrieve_node("wszd5")
#donwload familiar trials
#TO DO: separate this out so files are downloaded only once
creel_proj %>%
  osf_ls_files(pattern = "allfamiliars.csv") %>% # Replace "data.csv" with your file's name
  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")
creel_proj %>%
  osf_ls_files(pattern = "allnovels.csv") %>% # Replace "data.csv" with your file's name
  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")

#run Creel prep script
source(here("supplemental_data","creel_2024","Step2aCorrsNDataPrep.R"))
# run Peekbank-style rt computation over Creel data
source(here("supplemental_data","creel_2024","creel_peekbank_acc_rt_analysis.R"))

@@

To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024). This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image. Using this dataset, we investigated the correlation between reaction times (following the same procedure as in our main analyses, i.e. focusing specifically on distractor-to-target shifts) computed over all trials and reaction times computed only for those trials in which children selected the correct referent. The results are visualized below. Reaction times (i.e., distractor to target shifts) for correct trials only were highly correlated with reaction times across all trials (`r apa_print(creel_correlation)$full_result`). This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children. While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.

```{r}
creel_plot
```

@@

```{r}
library(fitdistrplus)
library(statmod)

@@

@@

```{r rt_mods}
rt_more_mods_lmer <- 
  list(log_log = lmer(log_rt ~ log_age_s + 
                        (log_age_s | dataset_name) + (1 | subject_id), 
                      data = d_trial),
       log_log_plus = lmer(log_rt ~ log_age_s + age_s +  
                             (log_age_s + age_s | dataset_name) + (1 | subject_id), 
                           data = d_trial), 
       lin_poly2 = lmer(rt ~ poly(age_s, 2) +  
                        (age_s | dataset_name) + (1 | subject_id), 
                      data = d_trial),
       lin_poly3 = lmer(rt ~ poly(age_s, 3) +  
                        (age_s | dataset_name) + (1 | subject_id), 
                      data = d_trial))

@@

```{r}
library(semoutput)
semoutput::sem_fitmeasures(fit3)
semoutput::sem_fitmeasures(fit3_firstadmin)
semoutput::sem_fitmeasures(fit3_age)
semoutput::sem_fitmeasures(fit3_age_firstadmin)

>>>>

```{r, eval=F}
#load data using Creel (2024) analysis scripts
# To run this script, you need to first pull in two datasets from the Creel (2024) OSF repository (https://osf.io/wszd5/):
#allfamiliars.csv and allnovels.csv
library(osfr)
library(here)
creel_proj <- osf_retrieve_node("wszd5")
#donwload familiar trials
#TO DO: separate this out so files are downloaded only once
creel_proj %>%
  osf_ls_files(pattern = "allfamiliars.csv") %>% # Replace "data.csv" with your file's name
  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")
creel_proj %>%
  osf_ls_files(pattern = "allnovels.csv") %>% # Replace "data.csv" with your file's name
  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")

creel_proj %>%
  osf_ls_files(pattern = "Step2aCorrsNDataPrep.R") %>% # Replace "data.csv" with your file's name
  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")
```

```{r, eval=F}
#run Creel prep script

#creel prep script expects files in root directory
setwd(here("supplemental_data","creel_2024"))
source("Step2aCorrsNDataPrep.R")
setwd(here())
# run Peekbank-style rt computation over Creel data
source(here("supplemental_data","creel_2024","creel_peekbank_acc_rt_analysis.R"))

@@

```{r}
# <!--To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024). This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image. Using this dataset, we investigated the correlation between reaction times (following the same procedure as in our main analyses, i.e. focusing specifically on distractor-to-target shifts) computed over all trials and reaction times computed only for those trials in which children selected the correct referent. The results are visualized below. Reaction times (i.e., distractor to target shifts) for correct trials only were highly correlated with reaction times across all trials (`r apa_print(creel_correlation)$full_result`). This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children. While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.-->

```{r, eval=F}
creel_plot
```

@@

```{r}
library(fitdistrplus)

library(statmod)

@@

#note that fitdistrplus attaches MASS which also has a select function, and that's annoying
detach("package:dplyr", character.only = TRUE)
library("dplyr", character.only = TRUE)

@@

```{r rt_more_mods}
rt_more_mods_lmer <- 
  list(log_log = lmer(log_rt ~ log_age_s + 
                        (log_age_s | dataset_name) + (1 | subject_id), 
                      data = d_trial),
       log_log_plus = lmer(log_rt ~ log_age_s + age_s +  
                             (log_age_s + age_s | dataset_name) + (1 | subject_id), 
                           data = d_trial), 
       lin_poly2 = lmer(rt ~ poly(age_s, 2) +  
                        (age_s | dataset_name) + (1 | subject_id), 
                      data = d_trial),
       lin_poly3 = lmer(rt ~ poly(age_s, 3) +  
                        (age_s | dataset_name) + (1 | subject_id), 
                      data = d_trial))

@@

```{r}
#remotes::install_github("dr-JT/semoutput")
library(semoutput)
semoutput::sem_fitmeasures(fit3)
semoutput::sem_fitmeasures(fit3_firstadmin)
semoutput::sem_fitmeasures(fit3_age)
semoutput::sem_fitmeasures(fit3_age_firstadmin)
